---
title: "Daily Activity Pattern Simulation"
author: "Greg Macfarlane"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(leaflet)
library(sf)
```

## Lab Task
Construct basic weekday daily activity patterns for Payson's current (2019) population. Your population should have three segments:

  - workers: these people should have one work activity and up to one other activity
  - non-workers: these people should have between zero and two other activities
  
Use MATSim's population and plans libraries to build these plans. You may
determine how to locate the households and their activities, but please document
how you did this in your memorandum. Run the scenario until you reach a
satisfactory level of convergence, and analyze the volumes at the I-15
interchange you are using for your Vissim analysis.


## Helps

In class we built a Java class file that will help you construct and write these
plans to an `xml.gz` file. You can find an example of this class file 
[here](materials/PlansMaker.java).

There are a few additional inputs you will need to create the plans. 

  - Population records (and forecasts for 2050 for the future volumes)
  - A distribution of points in Payson
  - A distribution of daily activity patterns by worker status
  

### Point Distribution in Payson
Using OpenStreetMap, we can determine that Payson can be generally described
with the following bounding box:

```
LATITUDE  > 40.0135   & LATITUDE  < 40.0680
LONGITUDE > -111.7771 & LONGITUDE < -111.7046
```

Let's get the population-weighted centroids for block groups in Utah that
fall inside this bounding box.

```{r payson_bg}
## pop-weighted block group centroids from Census bureau
url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG49.txt"
payson_bg <- read_csv(url) %>%
  filter(LATITUDE  > 40.0135, LATITUDE  < 40.0680) %>%
  filter(LONGITUDE > -111.7771, LONGITUDE < -111.7046)
```


```{r payson_bg_map, echo = FALSE}
leaflet(payson_bg %>%  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) ) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addCircleMarkers(radius = 1)
```

Then, let's calculate the mean and standard deviation of these points
in the $x$ and $y$ direction. The map following shows 1000 random points
with this mean and standard deviation. This method seems to work pretty
well, though it's not perfect given that Payson is largely on one side of the
freeway. But it'll work for this lab. You might also choose to use a uniform
distribution instead.

```{r payson_distribution}
(payson_distribution <- payson_bg %>%
  summarise(
    lat = weighted.mean(LATITUDE, w = POPULATION),
    lon = weighted.mean(LONGITUDE, w = POPULATION),
    lat_sd = sqrt(Hmisc::wtd.var(LATITUDE, POPULATION)),
    lon_sd = sqrt(Hmisc::wtd.var(LONGITUDE, POPULATION)),
  ))
```

```{r payson_houses}
n_people <- 1000
lat <- rnorm(n_people, mean = payson_distribution$lat, sd = payson_distribution$lat_sd)
lon <- rnorm(n_people, mean = payson_distribution$lon, sd = payson_distribution$lon_sd)

payson_houses <- tibble(id = 1:n_people) %>%
  mutate(
    lat = lat, lon = lon
  ) %>%
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)


leaflet(payson_houses) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addCircleMarkers(radius = 1)
```




