---
title: "Daily Activity Pattern Simulation"
author: "Greg Macfarlane"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(leaflet)
library(sf)
```

## Lab Task
Construct basic weekday daily activity patterns for Payson's current (2019) population. Your population should have three segments:

  - workers: these people should have one work activity and up to one other activity
  - non-workers: these people should have between zero and two other activities
  
Use MATSim's population and plans libraries to build these plans. You may
determine how to locate the households and their activities, but please document
how you did this in your memorandum. Run the scenario until you reach a
satisfactory level of convergence, and analyze the volumes at the I-15
interchange you are using for your Vissim analysis.


## Helps

In class we built a Java class file that will help you construct and write these
plans to an `xml.gz` file. You can find an example of this class file 
[here](materials/PlansMaker.java).

There are a few additional inputs you will need to create the plans. 

  - Population records (and forecasts for 2050 for the future volumes)
  - A distribution of points in Payson
  - A distribution of daily activity patterns by worker status
  
### Population Records
According the Google, the 2017 population of Payson was 19,892. Go ahead and 
use this for your 2019 target. The Kem C. Gardner Policy Institute 
[forecasts](https://gardner.utah.edu/wp-content/uploads/Projections-Brief-Final.pdf)
that Utah County will grow from 585,694 in 2015 to 1,396,997 by 2050, or 
`r round((1396997 - 585694)/585694 * 100, 2)`. If we apply the same growth rate
to Payson (doubtful, but easy), this would mean `r round((19892 * (1396997 - 585694)/585694) + 19892)`
will live in Payson in 2050.

> Can you think of a better way to get future population in the region under
study?

```{r youngold}
young <- c(9.9, 9.7, 9.2, 9.3) # shares of age groups less than 20
old <- c(4.2, 2.3, 0.7) # shares of age groups greater than 65
```

The labor force 
participation rate for all adults between 20 and 64 in the Provo-Orem metropolitan 
area is 76.8% [ACS](https://factfinder.census.gov/bkmk/table/1.0/en/ACS/17_5YR/S2301/312M300US393404962470).
The share of the population that is this age is `r 100 - sum(old, young)`%,
so the share of the total population that works is `r (1 - sum(old, young)*.01) * .768`
  
  
### Point Distribution in Payson
Using OpenStreetMap, we can determine that Payson can be generally described
with the following bounding box:

```
LATITUDE  > 40.0135   & LATITUDE  < 40.0680
LONGITUDE > -111.7771 & LONGITUDE < -111.7046
```

Let's get the population-weighted centroids for block groups in Utah that
fall inside this bounding box.

```{r payson_bg}
## pop-weighted block group centroids from Census bureau
url <- "https://www2.census.gov/geo/docs/reference/cenpop2010/blkgrp/CenPop2010_Mean_BG49.txt"
payson_bg <- read_csv(url) %>%
  filter(LATITUDE  > 40.0135, LATITUDE  < 40.0680) %>%
  filter(LONGITUDE > -111.7771, LONGITUDE < -111.7046)
```


```{r payson_bg_map, echo = FALSE}
leaflet(payson_bg %>%  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) ) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addCircleMarkers(radius = 1)
```

Then, let's calculate the mean and standard deviation of these points
in the $x$ and $y$ direction. The map following shows 1000 random points
with this mean and standard deviation. This method seems to work pretty
well, though it's not perfect given that Payson is largely on one side of the
freeway. But it'll work for this lab. You might also choose to use a uniform
distribution instead.

```{r payson_distribution}
(payson_distribution <- payson_bg %>%
  summarise(
    lat = weighted.mean(LATITUDE, w = POPULATION),
    lon = weighted.mean(LONGITUDE, w = POPULATION),
    lat_sd = sqrt(Hmisc::wtd.var(LATITUDE, POPULATION)),
    lon_sd = sqrt(Hmisc::wtd.var(LONGITUDE, POPULATION)),
  ))
```

```{r payson_houses}
n_people <- 1000
lat <- rnorm(n_people, mean = payson_distribution$lat, sd = payson_distribution$lat_sd)
lon <- rnorm(n_people, mean = payson_distribution$lon, sd = payson_distribution$lon_sd)

payson_houses <- tibble(id = 1:n_people) %>%
  mutate(
    lat = lat, lon = lon
  ) %>%
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)


leaflet(payson_houses) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addCircleMarkers(radius = 1)
```


### Daily Activity Pattern Distribution

We processed the 2017 National Household Travel Survey for metropolitan areas
between 1 and 3 Million people to determine the daily activity patterns for
the following two person types:

  - `workers`: People with a full or part-time job
  - `non-workers`: everyone else
  
It might be debatable whether Payson behaves more like the metropolitan region
it is a part of, or if it behaves like a small and isolated city. But dismissing
this quibble for the moment, the DAP distribution for these two person groups is 
given in the table below.

|Person Type |   Home   | Mandatory|Non-Mandatory|
|:-----------|---------:|---------:|---------:|
|non-worker  | 0.2294942| 0.1654482| 0.6050576|
|worker      | 0.0829301| 0.6316277| 0.2854422|


The definition of these patterns in the context of our simulation is as follows:

  - *Mandatory* patterns include one `work` activity and up to one additional
  after-work activity
  - *Non-mandatory* patterns include between one and three `other` activitities
  - *Home* patterns include no activities outside the home.



